Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2017-02-05 writeup/review-responses.tex

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Han Li \\
    \emph{and} Peter Ralph \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript,

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Han Li and
Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    Please understand that incremental changes will not be sufficient.  Adding
    simulations to strengthen key claims will be necessary, particularly addressing
    the impacts of mutation rate and recombination rate variation with more depth,
    the concern regarding PC switching (Reviewer 1), and the concern regarding the
    impacts of variation in missingness by sub-population (Reviewer 2).
\end{quote}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    The paper is generally well written and clear; it addresses an important
    problem, and clearly makes some progress on it. However, it suffers from having
    no grounding in either theory or empirical demonstration that it really can find
    the structures that are claimed. I find the arguments that it finds inversions
    compelling, though not watertight, and I am not yet convinced that it is finding
    ubiquitous background selection.  To make this claim, significant extra work is
    required.
\end{quote}

\begin{quote}
    In short, the approach is interesting but not sufficiently explored to produce
    compelling evidence for the implications that are claimed.  Putting a large
    amount of effort into simulations may alleviate these concerns somewhat.

    Specific points: What does this method find? I'm concerned about:
    (a) variation in the recombination rate
    and (b) variation in the mutation rate, creating spurious structure.

    The first possibility is that massively varying information quantity
    within windows could lead to a small number of such windows having their
    orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
    could switch). This would lead to such windows having unusual properties and
    hence appearing as evidence of an inversion.

    I do agree with the authors that significant outliers would be found at
    inversions. However, even if the PC switching does not occur, or the model could
    handle it, the evidence for selection is weaker.  If the two types of variation
    described above exist, with no selection, I would still expect a ``continuous
    triangle'' of results (as seen left of Fig 2, top left of Fig 6) with extrema
    described by windows with the most information, and points placed at different
    extremum having low recombination rate (because by chance, these will get an
    approximately fixed local tree, corresponding on average to the genome-wide
    population structure).

    Addressing this is likely quite hard, though the authors may be able to think of
    something that separates these effects from selection.
\end{quote}

\begin{point}{}
\ldots variation in the recombination rate \ldots 
\end{point}

\begin{point}{}
\ldots and variation in the mutation rate, creating spurious structure.
\end{point}

\begin{point}{PC switching}
\ldots could lead to a small number of such windows having their
orientation reversed: that is, PC1 becomes PC2 and vice versa. (Or PC2 and PC3
could switch).
\end{point}

\begin{point}{p6}
``here, we use k=2...''  - you have to show that $k>2$ is the same.
\end{point}

\begin{point}{p15}
``We also found nearly identical results when choosing shorter windows of 1,000 SNPs'' - again, show this.
\end{point}

\begin{point}{p15}
 ``or choosing windows of equal length in base pairs rather than SNPs'' - once again.
\end{point}

\begin{point}{}
Using 2 PCs is common practice: only if this is the end of an analysis and the
PCA was done for visualisation. Here you are using it for something so should
keep all the relavant PCs.
\end{point}

\begin{point}{}
I'm surprised that PCAdmix isn't referenced. It is using a very similar
method, albeit with different goals. In particular, the approach of placing all
points into a single, genome-wide PC space solves many of the problems that this
approach has (though I agree there may be benefits to the approach described here)
\end{point}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This is an interesting and well written paper. It was a pleasant read. I have three main general
    comments:
\end{quote}

\begin{point}{Related work:} 
The authors provide an introduction of the main concepts, as well as some
intuition of what the method is doing and how, but I found comparison to previous approaches
to be somewhat missing. To some extent, this is due to the fact that the main goal of their
analysis is somewhat vaguely ``finding heterogeneity'', which leads to the applications of
detecting chromosomal inversions and evidence for background selection. It would help to
have a well defined set of hypotheses, test the method’s accuracy  using simulation (see next
comment), and compare to previous efforts in similar domains.
\end{point}

\begin{point}{Validation:}
In several occasions, the authors seem to introduce a potential problem in their
approach, and provide a solution to it. This is generally rather intuitive, but it would really help
to have simulations of some sort to show that the issue arises and leads to a problem, and that
their approach does address the specific problem.
\end{point}

\begin{point}{}
The use of weighted PCA to cope with unbalanced sample size could be better demonstrated.
Although the current explanation makes intuitive sense, this approach does not seem to be
used in previous work. The authors could design a simulation that supports their approach.
\end{point}

\begin{point}{}
It is conceivable that some subpopulations will have more missingness in some windows. That
may skew the resulting PCs by selecting different sample sizes for the different windows (as
discussed in Appendix B) . This could distort the PCs, so that variation reflects underlying
variation in missingness. Would be good to discuss this potential issue and provide simulations.
\end{point}

\begin{point}{Appendix A:}
when using jackknife to estimate variance, each window is being divided in 10
``independent'' resampling units. Due to LD, these 10 blocks are likely correlated, which would
bias the estimates of variance. This is probably not a problem because both signal and noise
could be equally biased, but the authors may want to consider this potential issue. I wonder if
the correlation with recombination rate may be partially explained by this.
\end{point}

\begin{point}{}
Is it possible to explain the results of Figure 6 just considering neutral variation in local
ancestry due to recent admixture? This may explain why ancestry seems to explain a fair
amount of variance in the lower plots of Fig 6. Local PCA has been previously used by others to
detect local ancestry blocks, e.g. see the PCAdmix approach by Brisbin et al. The authors
discuss the possibility that admixture is driving the differentiation, but do not test whether their
observations agree with neutrality.
\end{point}

\begin{point}{}
``to remove the effect of artifacts such as mutation rate variation, we also rescale each
approximate covariance matrix to be of similar size (precisely, so that the underlying data
matrix has trace norm equal to one''. This potential issue is a bit unclear to me, since I would
expect that scaling the volume of local trees would not result in changed distances in PC
space. Perhaps the authors could show via simulations that this creates a problem, and that the
normalization addresses it.
\end{point}

\begin{point}{Figure 7:}
are MDS coordinates correlated with recombination rates in this case?
\end{point}

\begin{point}{Application:}
 is what the authors seem to be proposing not already accounted for by linear
mixed model association approaches? If not, this should be clarified. Either way, this paragraph
could be dropped.
\end{point}

\begin{point}{Introduction:}
 ``it is not necessarily clear what aspects of demography should be included in the
concept.'' I find it a bit weird to describe selection as an ``aspect of demography''. Although it
could be seen as such within a coalescent framework, that seems to be just a useful
representation. The authors may consider rewording`.
\end{point}

\begin{point}{}
Paragraph starting in ``Since the definition...''. The notation is a bit unclear. Please check that it
is clear which PC the text refers to.
\end{point}

\begin{point}{}
Would the authors be able to provide a sense for the directionality of effects in Figure 4? It
would be interesting if the authors tried to further characterize regions that are similar due to
higher recombination rates. E.g. is there more/less density of polymorphisms in these regions?
\end{point}

\begin{point}{Page 13:}
typo: ``figures 6 and 6''.
\end{point}

\begin{point}{}
Typo in abstract, line 6 ``, We show'' -> ``. We show''.
\end{point}

\begin{point}{}
Typo: end of introduction ``an visualization''. The whole sentence is a bit weird. The authors just
stated focus is on clustering, not on looking for outliers, but what does it mean that ``we allow
ourselves to be surprised by unexpected signals in the data''?
\end{point}

\begin{point}{}
``There has been substantial debate over the relative impacts of different forms of selection.''
Citation needed.
\end{point}

\begin{point}{}
``Results using larger numbers of PCs were nearly identical''. It would be interesting to have a
supplementary table.
\end{point}

\begin{point}{}
Table 1 legend seems a bit redundant. Columns are self-explanatory.
\end{point}

\begin{point}{}
It would help to have numbered lines and references.
\end{point}

### 2017-02-05 writeup/local_pca_paper.tex
background sims seem to work
added sqrt to distance computation
added response to reviewers to latex

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}

@@

@@

The goal of comparing PC plots up to rotation and reflection 
turned out to be equivalent to comparing rank-$k$ approximations to local covariance matrices.
This suggests instead directly comparing entire local covariance matrices. 
However, with thousands of samples and tens of thousands of windows,
computing the distance matrix would take months of CPU time,
while as defined above, $D$ can be computed in minutes using the following method.
Since for square matrices $A$ and $B$,
$\sum_{ij} (A_{ij}-B_{ij})^2 = \sum_{ij} (A^2_{ij} + B^2_{ij}) - 2 \tr(A^T B)$,
then due to the orthogonality of eigenvectors and the cyclic invariance of trace,
$D_{ij}$ can be computed efficiently as
\begin{align}
    D_{ij} 
    = 
    \frac{ \sum_{\ell=1}^k \lambda_{\ell i}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell i})^2 }
    + \frac{ \sum_{\ell=1}^k \lambda_{\ell j}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell j})^2 }
    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} .
\end{align}
% from the code for pc_dist():
% #' In the unweighted case, suppose that (u) and (v) are sets of vectors, and that 
% #'    A = a_1 * u_1 u_1^T + ... + a_j * u_j u_j^T = U diag(a) U^T
% #'    B = b_1 * v_1 v_1^T + ... + b_j * v_k v_k^T = V diag(b) V^T,
% #' where U is the matrix whose columns are (u), and likewise for V.
% #' Then 
% #'    (A-B)^T (A-B) = A^T A + B^T B - A^T B - B^T A
% #' and so
% #'    ||A-B||^2 = ||A||^2 + ||B||^2 - 2 tr( A^T B )
% #' By the cyclic invariance of trace, if X = U^T V then
% #'    tr( A^T B ) = tr( U diag(a) U^T V diag(b) V^T ) = tr( diag(a) X diag(b) X^T ) .
% \begin{align}
%     \begin{split}
%     D_{ij} &= 
%         \left \{
%                 \left ( 
%                     V_{i1}\cdot V_{i1}^{} 
%                 \right )^{2}
%                 + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
%                 + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
%                 +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
%         \right. \\ 
%         & \left. \qquad {}  
%             -2\left [ 
%                 \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
%                 + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
%                 + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
%                 + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
%             \right ]
%         \right \}^{1/2}
%     \end{split}
% \end{align}

@@

We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
for assistance with obtaining data and interpreting results
(especially inversion status of \textit{D.~melanogaster} samples).
Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
and for encouraging the project.

@@

>>>>

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}

@@

%% for submitting with responses to reviews
\usepackage{lineno}

% from http://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation/55297#55297
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file

@@

The goal of comparing PC plots up to rotation and reflection 
turned out to be equivalent to comparing rank-$k$ approximations to local covariance matrices.
This suggests instead directly comparing entire local covariance matrices. 
However, with thousands of samples and tens of thousands of windows,
computing the distance matrix would take months of CPU time,
while as defined above, $D$ can be computed in minutes using the following method.
Since for square matrices $A$ and $B$,
$\sum_{ij} (A_{ij}-B_{ij})^2 = \sum_{ij} (A^2_{ij} + B^2_{ij}) - 2 \tr(A^T B)$,
then due to the orthogonality of eigenvectors and the cyclic invariance of trace,
$D_{ij}$ can be computed efficiently as
\begin{align}
    D_{ij} 
    = 
    \left(
    \frac{ \sum_{\ell=1}^k \lambda_{\ell i}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell i})^2 }
    + \frac{ \sum_{\ell=1}^k \lambda_{\ell j}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell j})^2 }
    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} 
    \right)^{1/2}.
\end{align}
% from the code for pc_dist():
% #' In the unweighted case, suppose that (u) and (v) are sets of vectors, and that 
% #'    A = a_1 * u_1 u_1^T + ... + a_j * u_j u_j^T = U diag(a) U^T
% #'    B = b_1 * v_1 v_1^T + ... + b_j * v_k v_k^T = V diag(b) V^T,
% #' where U is the matrix whose columns are (u), and likewise for V.
% #' Then 
% #'    (A-B)^T (A-B) = A^T A + B^T B - A^T B - B^T A
% #' and so
% #'    ||A-B||^2 = ||A||^2 + ||B||^2 - 2 tr( A^T B )
% #' By the cyclic invariance of trace, if X = U^T V then
% #'    tr( A^T B ) = tr( U diag(a) U^T V diag(b) V^T ) = tr( diag(a) X diag(b) X^T ) .
% \begin{align}
%     \begin{split}
%     D_{ij} &= 
%         \left \{
%                 \left ( 
%                     V_{i1}\cdot V_{i1}^{} 
%                 \right )^{2}
%                 + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
%                 + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
%                 +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
%         \right. \\ 
%         & \left. \qquad {}  
%             -2\left [ 
%                 \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
%                 + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
%                 + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
%                 + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
%             \right ]
%         \right \}^{1/2}
%     \end{split}
% \end{align}

@@

We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
for assistance with obtaining data and interpreting results
(especially inversion status of \textit{D.~melanogaster} samples).
Jaime Ashander and Jerome Kelleher provided assistance in performing the simulations.
Thanks also go to Yaniv Brandvain, Barbara Engelhardt, Charles Langley, Graham Coop, and Jeremy Berg for helpful comments
and for encouraging the project.

@@

% reviews here if toggled above
\includereviews
